---
title: "single R markdown"
author: "SALMA"
date: "3/27/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Coursera - Reproducible Research: Peer Assessment 1


#loading and processind the data
load the data

```{r}
activity <- unzip("activity.zip")
act_data <- read.csv("activity.csv", header=TRUE, sep=",")
str(act_data)
```

Process the data
```{r}
act_data$date <- as.Date(act_data$date, format="%Y-%m-%d")
act_data$interval <- as.factor(act_data$interval)
```

Post converting the column classes print the structure of the data
```{r}
str(act_data)
```

Print the header of the dataset
```{r}
head(act_data, 10)
```

#What is mean total number of steps taken per day?
##1-Calculate the total number of steps taken per day
```{r}
steps_per_day <- aggregate(steps ~ date, data=act_data, FUN=sum)
colnames(steps_per_day) <- c("date", "steps")
```

Print the header of the data frame with steps aggregated per day
```{r}
head(steps_per_day, 10)
```

##2. Make a histogram of the total number of steps taken each day
```{r}
library(ggplot2)
ggplot(steps_per_day, aes(x = steps)) + 
  geom_histogram(fill = "blue", binwidth = 1000) + 
  labs(title = "Histogram - Steps Taken Per Day", x = "Steps Per Day", y = "Frequency")
```

##3. Calculate and report the mean and median of the total number of steps taken per day
```{r}
mean_steps_per_day <- mean(steps_per_day$steps)
mean_steps_per_day
median_steps_per_day <- median(steps_per_day$steps)
median_steps_per_day
```

#What is the average daily activity pattern?
##1.Make a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)
```{r}
steps_per_interval <- aggregate(steps ~ interval, data = act_data, FUN = mean, na.rm = TRUE)
steps_per_interval$interval <- as.integer(levels(steps_per_interval$interval)[steps_per_interval$interval])
colnames(steps_per_interval) <- c("interval", "steps")
```

Print the header of the data frame with steps aggregated per interval
```{r}
head(steps_per_interval, 10)
```

Plot the timeseries graph
```{r}
ggplot(steps_per_interval, aes(x = interval, y = steps)) + 
  geom_line(col = "blue", size = 1) + 
  labs(title = "Average Daily Activity Pattern", x = "Interval", y = "Steps")
```

##2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
```{r}
max_interval <- steps_per_interval[which.max(steps_per_interval$steps),]
max_interval
```
#Imputing missing values
##1.Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
```{r}
missing_values <- sum(is.na(act_data$steps))
missing_values
```
##2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.
To populate missing values, we choose to replace them with the mean value at the same interval across days.
##3. Create a new dataset that is equal to the original dataset but with the missing data filled in.
```{r}
new_act_data <- act_data
index_of_na <- which(is.na(new_act_data$steps))
for (i in index_of_na) {
  new_act_data$steps[i] <- with(steps_per_interval, steps[interval = new_act_data$interval[i]])
}
```
Print the top rows of newly created dataset
```{r}
head(new_act_data, 10)
```
Given we have handled the missing values, let us check if the above strategy really worked out.
```{r}
new_missing_values <- sum(is.na(new_act_data$steps))
new_missing_values
```

##4. Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?
```{r}
new_steps_per_day <- aggregate(steps ~ date, data = new_act_data, FUN=sum)
colnames(new_steps_per_day) <- c("date", "steps")
ggplot(new_steps_per_day, aes(x = steps)) + 
  geom_histogram(fill = "blue", binwidth = 1000) + 
  labs(title = "Histogram - Steps Taken Per Day", x = "Steps Per Day", y = "Frequency")
```
In order to find the impact of imputing the missing values, let us compute the mean and median of steps taken per day
```{r}
new_mean_steps_per_day <- mean(new_steps_per_day$steps)
new_mean_steps_per_day
new_median_steps_per_day <- median(new_steps_per_day$steps)
new_median_steps_per_day
```
As we can see both mean and median of steps taken per day became exactly same which was not the case prior to imputing missing values

#Are there differences in activity patterns between weekdays and weekends?

##1. Create a new factor variable in the dataset with two levels - “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
Let us first add a factor variable to identify the given date as Weekday or Weekend
```{r}
library(data.table)
dt <- data.table(new_act_data)
dt[, weekday := ifelse(weekdays(date) %in% c("Saturday", "Sunday"), "Weekend", "Weekday")]
dt$weekday <- as.factor(dt$weekday)
dt$interval <- as.integer(levels(dt$interval)[dt$interval])
head(dt, 10)
```
##2. Make a panel plot containing a time series plot (i.e. type = “l”) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis)
```{r}
steps_per_weekday <- aggregate(steps ~ interval+weekday, data = dt, FUN = mean)
ggplot(steps_per_weekday, aes(x = interval, y = steps)) + 
  geom_line(col = "blue", size = 1) + 
  facet_wrap(~ weekday, nrow=2, ncol=1) + 
  labs(x = "Interval", y = "Number of Steps")
```
